Normalization by Evaluation for Sized Dependent Types

### 2017-06-09 nbe/icfp17.tex
[ ICFP 17 ] Started polishing final version.

%% For double-blind review submission
\documentclass[acmlarge,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

@@

Dependently-typed programming languages and proof assistants, such as Agda~\citeyearpar{agdawiki}, Coq~\cite{inria:coq86}, and Idris~\cite{brady:jfp13} require programs to be total, for two reasons.  First, for consistency:  since propositions are just types and proofs of a proposition just programs which inhabit the corresponding type, some types need to be empty; otherwise, each proposition would be true.  However, in a partial language with general recursion, each type is inhabited by the looping program $f = f$.  Secondly, totality is needed for decidability of type checking.  Since types can be the result of a computation, we need computation to terminate during type checking, even for \emph{open terms}, \ie, terms with free variables.

% \cite{boveKraussSozeau:mscs16}
Consequently, each of the aforementioned languages based on Type Theory comes with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.  In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.  This is not a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.  However, the syntactic termination check is very sensitive to reformulations of the program and hostile to abstraction \cite{abel:fics12}.

@@

Higher-rank size quantification takes termination checking to the next
level; it is necessary for abstraction and generic programming.  For
instance, it allows us to write a generic tree traversal which applies
a user-given preprocessor on subtrees before recursively descending
into these trees, and a postprocessor after surfacing from the
descent.
The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i)\; (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and should be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check can be hazardous, especially
when the equality-check is type directed.

@@

Going beyond the classical incremental equality check, there is a technique that can deal with dynamic $\eta$-expansion.  It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the computed normal forms for identity and, thus, decide definitional equality.
NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

@@

We present our type theory as \emph{pure type system} \cite{barendregt:lambdacube} with extra structure.
\emph{Sorts} $s$ are drawn from an infinite predicative hierarchy of universes $\Set[\ell]$ for $\ell \in \NN$.  Universes provide us with polymorphism and the capability to define types by recursion on values.  Whether we have just two universes $\Set[0]$ and $\Set[1]$ or infinitely many, does not matter for the technical difficulty of the meta theory.  The present setup have the advantage that every sort has again a sort since $\Set[\ell] : \Set[\ell+1]$, thus, we do not have to introduce a separate judgement $\Gamma \der T$ for well-formedness of types, we can define it as $\exists s.\ \Gamma \der T : s$.

@@

For the expression syntax (see Fig.~\ref{fig:syntax}), we use de~Bruijn \citeyearpar{deBruijn:nameless} indices $\ind i$ to represent variables.  Binders are lambda abstraction $\lambda t$ and dependent function types $\epiT U T$, which bind the $0$th index in $t$ and $T$, resp.  The notation $\epiT U T$ is an umbrella for three kinds of function types, where $\evar \in \{ \erased, \noterased \}$ is a relevance annotation borrowed from \citet{pfenning:intextirr}: $\erpiT[\noterased] U T$, the ordinary dependent function type, $\erpiT[\noterased] \Size T$ for relevant size quantification, and $\erpiT \Size T$ for irrelevant size quantification.
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default and write $\forallT T$ for $\erpiT \Size T$.
%
Examples for relevant size quantification $\piT \Size T$, in syntax with named variables, are $\Size \to \Set[0]$ or $\funT i \Size \Nat i \to \Set[0]$.  An instance of irrelevant quantification $\forall T$ would be $\forall i.\, \Nat i \to \Nat i$.  Herein, $\Nat i$ denotes the type of natural numbers below $i$. The expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a possible type of a variable in a typing context $\Gamma$, but not a first-class type, \ie, we cannot construct our own function on sizes.

@@

Besides with regular application $t\,u$, relevant size application $t\,a$, and irrelevant size application $t \, \ann a$,
we can eliminate a term $t$ by a case distinction $e = \case[\ell] T {t_z}{t_s}$ or a recursive function application $e' = \fix[\ell'] {T'} {t'}$.  In both cases, the term should denote a natural number.  Application of case distinction $\zero a \, e$ will reduce to the zero-branch $t_z$, and $\suc a t \,e$ to the instantiation of the successor branch $t_s\,t$.  The type annotation $T$ in $\tcase$ allows us to infer the type of the whole case statement $t\,e$ as $T\,t$.  The function call $c\,e'$ for a canonical number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed ourselves the use of a named abstraction in the presentation to the reader.  The unfolding of fixed-points is thus restricted to application to canonical numbers; this is the usual reduction strategy which converges for terminating functions \cite{gimenez:typeBased}.

@@

\clearpage

>>>>

%% For double-blind review submission
\documentclass[acmlarge%,review,anonymous
]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

@@

% \acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-functional.png}
\acmBadgeR[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_evaluated-reusable.png}
%\acmBadgeL[http://icfp17.sigplan.org/track/icfp-2017-Artifacts]{artifact_available.png}
\acmBadgeL[https://hackage.haskell.org/package/Sit]{artifact_available.png}

@@

Dependently-typed programming languages and proof assistants, such as Agda~\citeyearpar{agdawiki} and Coq~\cite{inria:coq86}, require programs to be total, for two reasons.  First, for consistency:  since propositions are just types and proofs of a proposition just programs which inhabit the corresponding type, some types need to be empty; otherwise, each proposition would be true.  However, in a partial language with general recursion, each type is inhabited by the looping program $f = f$.  Secondly, totality is needed for decidability of type checking.  Since types can be the result of a computation, we need computation to terminate during type checking, even for \emph{open terms}, \ie, terms with free variables.

% \cite{boveKraussSozeau:mscs16}
Consequently, the aforementioned languages based on Type Theory come with a termination checker, which needs to reject all non-terminating programs, and should accept sufficiently many terminating programs to allow the user to express her algorithms.  In current termination checkers, programs are required to terminate by structural descent \cite{gimenez:guardeddefinitions}; the structural order may be extended to a lexicographic \cite{abelAltenkirch:jfp02} or size-change termination criterion \cite{jones:sizeChange,wahlstedt:PhD}.  This is not a fundamental limitation, since Type Theory allows many functions to be expressed in a structurally recursive manner, if needed by the help of a well-founded relation \cite{nordstrom:genrec}, inductive domain predicates \cite{boveCapretta:mscs05}, or inductive descriptions of the function graph \cite{bove:entcs09}.  However, the syntactic termination check is very sensitive to reformulations of the program and hostile to abstraction \cite{abel:fics12}.

@@

Higher-rank size quantification takes termination checking to the next
level; it is necessary for abstraction and generic programming.  For
instance, it allows us to write a generic tree traversal which applies
a user-given preprocessor on subtrees before recursively descending
into these trees, and a postprocessor after surfacing from the
descent.
The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i)\; (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris~\cite{brady:jfp13}, and Haskell~\cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and should be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check needs special consideration, especially
when the equality-check is type directed.

@@

Going beyond the standard syntax-directed equality check, there is a technique that can deal with dynamic $\eta$-expansion.  It is a type-directed normalization function inspired by normalization-by-evaluation (NbE) that computes $\eta$-long normal forms \cite{bergerSchwichtenberg:lics91,danvy:tdpe}.  We can check the computed normal forms for identity and, thus, decide definitional equality.
NbE has proven to be a robust method to decide equality in powerful type theories with non-trivial $\eta$-laws.  It scales to universes and large eliminations \cite{abelCoquandDybjer:lics07}, topped with singleton types or proof irrelevance \cite{abelCoquandPagano:lmcs11}, and even impredicativity \cite{abel:flops10}.  At its heart there are reflection $\up T$ and reification $\down T$ functions directed by type $T$ and orchestrating just-in-time $\eta$-expansion.  Reflection $\up T x$ maps variables $x$ into the realm of values of type $T$ and let's us compute with open terms.  Reification $\down T a$ takes a value $a$ of type $T$ and computes its long normal form.  For instance, the normal form of a closed function $f : U \to T$ would be $\lambda x.\,\down{T} (f\,(\up U x))$, and for its dependently-typed variant $f : \funT x U T[x]$ it would be $\lambda x.\,\down{T[\up U x]} (f\,(\up U x))$.

@@

We present our type theory as (domain-free) \emph{pure type system} \cite{barendregt:lambdacube} with extra structure.
The \emph{sorts} $s$ are drawn from an infinite predicative hierarchy of universes $\Set[\ell]$ for $\ell \in \NN$.  Universes provide us with polymorphism and the capability to define types by recursion on values.  Whether we have just two universes $\Set[0]$ and $\Set[1]$ or infinitely many, does not matter for the technical difficulty of the meta theory.  The present setup have the advantage that every sort has again a sort since $\Set[\ell] : \Set[\ell+1]$, thus, we do not have to introduce a separate judgement $\Gamma \der T$ for well-formedness of types, we can define it as $\exists s.\ \Gamma \der T : s$.

@@

For the expression syntax (see Fig.~\ref{fig:syntax}), we use
de~Bruijn \citeyearpar{deBruijn:nameless} indices $\ind i$ to
represent variables.  The index $i \in \NN$ points to the $i$th
enclosing binder of variable $\ind i$.  Binders are lambda abstraction
$\lambda t$ and dependent function types $\epiT U T$, which bind the
$0$th index in $t$ and $T$, resp.  For instance, the term
$\lambda x.\,x\,(\lambda z.\,z)\, (\lambda y.\,x\,y)$
with named variables $x,y,z$ has de Bruijn representation
$\lambda.\,\ind 0\,(\lambda.\, \ind 0)\,(\lambda.\,\ind 1\,\ind0)$.

The notation $\epiT U T$ is an
umbrella for three kinds of function types, where
$\evar \in \{ \erased, \noterased \}$ is a relevance annotation
borrowed from \citet{pfenning:intextirr}. $\erpiT[\noterased] U T$ is
the ordinary dependent function type, $\erpiT[\noterased] \Size T$ is
relevant size quantification, and $\erpiT \Size T$ is irrelevant size
quantification.  We omit the ``$\noterased$''-markers
from $\Pi$ by default (and also in contexts $\Gamma$)
and write $\forallT T$ for $\erpiT \Size T$.
%
Examples for relevant size quantification $\piT \Size T$, in syntax with named variables, are $\Size \to \Set[0]$ or $\funT i \Size \Nat i \to \Set[0]$.  An instance of irrelevant quantification $\forall T$ would be $\forall i.\, \Nat i \to \Nat i$.  Herein, $\Nat i$ denotes the type of natural numbers below $i$. The expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a possible type of a variable in a typing context $\Gamma$, but not a first-class type, \ie, we cannot construct our own function on sizes.

@@

Besides with regular application $t\,u$, relevant size application
$t\,a$, and irrelevant size application $t \, \ann a$, we can
eliminate a term $t$ by a case distinction
$e = \case[\ell] T {t_z}{t_s}$ or a recursive function application
$e' = \fix[\ell'] {T'} {t'}$.  In both cases, the term should denote a
natural number.  Application of case distinction $\zero a \, e$ will
reduce to the zero-branch $t_z$, and $(\suc a t) \,e$ to the
instantiation of the successor branch $t_s\,t$.  The type annotation
$T$ in $\tcase$ allows us to infer the type of the whole case
statement $t\,e$ as $T\,t$.  The function call $c\,e'$ for a canonical
number reduces to $f'\,(\lambda x.\,x\,e')\,c$ where we allowed
ourselves the use of a named abstraction in the presentation to the
reader.  The unfolding of fixed-points is thus restricted to
application to canonical numbers; this is the usual reduction strategy
which converges for terminating functions \cite{gimenez:typeBased}.

@@

% \clearpage
